### 2025-01-22 paper/icml25/first/appendices/ablation.tex
appendix wip

\begin{figure*}[!t]
    \centering
    \includegraphics[width=\textwidth]{figs/ablation-transformer.pdf}
    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios.}
    \label{fig:ablation}
\end{figure*}

\section{Ablation Study}\label{apx:ablation}
While we mainly focus on replacing both the reward and transition function with neural networks, we also investigate the effect of replacing only the latter.

In fact, the transition function is the most complex component to replace as it requires modeling with precision the observations obtained from the environment after a number of $H$ actions from each agent. Avoiding the need of such network lightens the algorithm but it reintroduces the need for a differentiable environment.
\Cref{fig:ablation} presents a comparative analysis between \fname{} and \fnamer{}. The results demonstrate that utilizing the true transition function significantly impacts the performance of \fname{}. This performance difference is particularly pronounced in the Sampling and Transport scenarios, where \fname{} consistently achieves superior results compared to \fnamer{}, both in terms of maximum reward and sample efficiency. This advantage is most evident in the Sampling scenario with one and three agents.

However, the performance gap narrows in certain scenarios. For instance, in the Dispersion scenario, \fnamer{} achieves comparable performance to \fname{}. This equivalence might be attributed to the increased complexity of learning the transition function in this particular environment.

>>>>

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figs/ablation-transformer.pdf}
    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios. Transformer architecture.}
    \label{fig:ablation-transformer}
\end{figure}

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/ablation-mlp.pdf}
    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios. MLP architecture.}
    \label{fig:ablation-mlp}
\end{figure}

\section{Ablation Study}\label{apx:ablation}
In the main paper, we focused on simultaneously replacing both the transition and reward functions with neural networks. This approach allows for greater generality and can be applied across a multitude of environments. However, when only the transition or reward function is differentiable and available, it is feasible to replace only one of them. This simplifies the algorithm, requiring the training of only a single additional network alongside the policy and value functions.

Specifically, we focus on replacing only the reward function. In fact, transition dynamics are often complex and challenging to model with a neural network, especially in multi-agent environments where observations depend not only on the actions of individual agents but also on those of other agents. This, in turn, may necessitate the use of larger architectures with greater computational demands, ultimately limiting the algorithm's practicality.

In contrast, the reward function is often simpler and, therefore, easier to model using lightweight neural networks. Moreover, a high degree of precision is typically not required in most scenarios. As a result, the reward function is an ideal candidate for replacement with a neural network.

This experiment also enables us to compare the final impact of gradients generated by a complex physics engine with those generated by a neural network. While the physics engine may produce more unstable gradients, the neural network may generate imprecise gradients, particularly during the early stages of training.

\subsection{Results}
We refer to the \fname{} algorithm as \fnamer{} when the transition function is differentiable and directly available. The results for the transformer architecture are shown in \Cref{fig:ablation-transformer}, while the results for the MLP architecture are presented in \Cref{fig:ablation-mlp}.

For the transformer architecture, \fname{} generally appears to generate better policies, particularly in the Sampling and Transport environments. In the Dispersion environment, both algorithms perform comparably. Slightly better results are achieved with the \fnamer{} algorithm in the Discovery environment.

For the MLP architecture, the results are more mixed. In the Dispersion environment, \fname{} appears to achieve better policies. Conversely, in the Transport environment, \fnamer{} leads to substantially better policies. In the Discovery environment, both policies struggle to produce meaningful results. Finally, in the Sampling environment, \fname{} performs on par with \fnamer{}, except in the single-agent case, where \fnamer{} struggles.

\subsection{Discussion}
We can easily observe that the transformer architecture leads, generally speaking, to better policies. This is also evident by inspecting \Cref{tab:max-rewards}. This may be due to the positional invariance property of the transformers leading to more sample efficient networks.

Furthermore, with the transformer architecture, we observe a tendency for \fname{} to produce better results. This may be attributed to the generated gradients being more stable, as shown in ~\cref{TODO}.

In contrast, for the MLP architecture, the results are more mixed. This could be attributed to the MLP architecture being less sample-efficient, which ultimately leads to less accurate gradients.

### 2025-01-22 paper/icml25/first/appendices/additional.tex
appendix wip

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/main-mlp.pdf}
    \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios with the MLP architecture.
    We show the mean and standard deviation of the normalized rewards across $3$ runs. For the Transformer architecture see \Cref{fig:experiments}.
    }
    \label{apx:fig:experiments-mlp}
\end{figure}

\begin{figure*}[!t]
    \centering
    \includegraphics[width=\textwidth]{figs/ablation-transformer.pdf}
    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios.}
    \label{fig:ablation}
\end{figure*}

\begin{figure}[!t]
    \centering
    \includegraphics[width=\columnwidth]{figs/ablation-mlp.pdf}
    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios.}
    \label{fig:ablation}
\end{figure}

>>>>

Finally, we present the results of the experiments with the MLP architecture in \Cref{apx:fig:experiments-mlp}. We observe that the MLP architecture underperforms compared to the Transformer architecture. Nonetheless, \fname{} often outperforms PPO and generally performs on par with SHAC.

Interestingly, the Discovery scenario appears to be the most challenging, as none of the algorithms (PPO and \fname{}) are able to make any meaningful progress. This may be caused by the sample inefficiency of the MLP architecture, combined with the fact that the world model cannot access the full state.

The only scenario where SHAC outperforms \fname{} is the Transport scenario. This may be due to the MLP's inability to overcome the gradient noise from the world model.

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/main-mlp.pdf}
    \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios with the MLP architecture.
    We show the mean and standard deviation of the normalized rewards across $3$ runs. For the MLP architecture see \Cref{fig:experiments}.
    }
    \label{apx:fig:experiments-mlp}

\end{figure}

### 2025-01-22 paper/icml25/first/appendices/arch.tex
appendix wip

@@

>>>>

\begin{figure}[t]
    \centering
    \includegraphics[width=.5\textwidth]{figs/attention.pdf} 
    \caption{Attention matrix for the world model.}
    \label{fig:apx:attention}
\end{figure}

@@

As mentioned earlier, we test PPO, SHAC, \fname{}, and \fnamer{} with two different architecture---Transformer and MLP. The former is a popular positional invariant architecture for sequence modeling~\cite{Vaswani17}. While the latter is a simple feed-forward neural network.

The main advantage of the transformer architecture is its positional invariance property. In fact, the order in which the agent's observation are fed into the network should not affect the ultimate outcome, at least, for the tested environments. We believe that this leads to better sample efficiency that it is ultimately observed in policies achieving higher rewards (see \Cref{tab:max-rewards}).

In contrast, the MLP is fed with a concatenation of all agent observations. This results in a simple and lightweight implementation, leading to faster inference times. However, in our experiments, the MLP appears to underperform compared to the transformer architecture.

\subsection{Policy/Value/Reward MLP Network}
When implementing the policy, value, and reward functions with an MLP, we use a single hidden layer with sizes of $32$, $96$, and $160$, depending on the number of agents. This choice was made to account for the increasing complexity that a higher number of agents should introduce. The activation function is always ReLU, and dropout is used only for the policy network. The learning rate is consistently set to $0.001$. The cache for training these networks is set to $30,000$ samples. Gradient clipping is applied only to the policy network, with a threshold of $1$. For a full list of hyperparameters, see \Cref{apx:tab:shac} and \Cref{apx:tab:ppo}.

\subsection{Policy/Value/Reward Transformer Network}
When implementing the policy, value, and reward functions with a transformer, we use a single hidden layer with a size of $64$, regardless of the number of agents, and a feed-forward size of $128$. We do not scale these sizes with the number of agents, as the transformer is expected to natively scale with the sequence length. The remaining hyperparameters are set consistently with those used in the MLP architecture. For a full list of hyperparameters, see \Cref{apx:tab:shac} and \Cref{apx:tab:ppo}.

\subsection{World Model}
When implementing the action world model, we consistently use a transformer architecture. Each action is represented as an embedding fed into the network. To maintain positional invariance between the agents, we employ a custom attention matrix that allows the action for step $i$ of agent $j$, $a_{ij}$, to attend only to actions from the same step or previous steps of the same agent. Formally, $a_{ij}$ can attend to all $a_{kj}$ for $k<i$ and to all $a_{ik}$ for $k \in \agents$. This attention pattern is shown in \Cref{apx:fig:attention}. As a result, the world models process sequences of length $32 \cdot \na$. We use a slightly deeper architecture to account for the increased complexity of the task, with $3$ layers. The hidden size is set to $64$ and the feed-forward size to $128$. The remaining hyperparameters are set consistently with those used in the MLP architecture. For a full list of hyperparameters, see \Cref{apx:tab:shac}.
